---
title: "Using Principal Component Analysis to explore co-variation of vowels."
author: "Joshua Wilson Black<sup>1</sup>, Jen Hay<sup>1,2</sup>, Lynn Clark<sup>1,2</sup>, & James Brand<sup>3</sup>.<br/><br/><sup>1</sup>New Zealand Institute for Language, Brain and Behaviour, Univeristy of Canterbury, NZ<br> <sup>2</sup>Department of Linguistics, Univeristy of Canterbury, NZ<br><sup>3</sup>Institute of Czech Language and Theory of Communication, Charles University, CZ<br/><br/>Corresponding author: Joshua Wilson Black<br/>Email: joshua.black@canterbury.ac.nz<br/>Website: https://joshua.wilsonblack.nz"
date: "2022-09-12"
output:
  bookdown::html_document2:
    fig_caption: true
    toc: true
    toc_float: true
    theme: flatly
    collapsed: no
    df_print: paged
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Overview

This RMarkdown presents all code used in the article 'Using Principal Component
Analysis to explore co-variation of vowels' and additional discussion.
Throughout, we use data from the [ONZE corpus](https://www.canterbury.ac.nz/nzilbb/research/onze/).

Section One covers the toy example used to present PCA. We take three variables
with known interrelations: the mean raw F1 values of <span style="font-variant: small-caps;">dress</span>,
<span style="font-variant: small-caps;">kits</span>, and <span style="font-variant: small-caps;">trap</span>.
These values are taken from a sample of 100 speakers from the ONZE corpus. By
looking at _three_ variables, we can visualise the entire process. By looking at
three variables _with known relationships_, we can be confident in our
interpretation of our PCA analysis.

Section Two provides a standalone variant of the analysis carried out in
[Brand et al. (2021)](https://www.sciencedirect.com/science/article/pii/S0095447021000711).
The details of the implementation and models have been slightly changed to
enable faster computation.^[
For full details of the analysis of Brand et al., see the original paper and 
the supplementary analysis file available athttps://nzilbb.github.io/Covariation_monophthongs_NZE/Covariation_monophthongs_analysis.html and
on the .
] 
Unlike the analysis as it first appears in Brand et
al. (2021),  the version of the analysis presented here is carried out with
generalisation at the front of mind. In particular, we will go into detail
regarding the three core questions asked of any PCA analysis:

1. Is PCA appropriate?
2. How many PCs should we consider?
3. Which variables should we use to interpret our PCs?

Throughout, we use functions from the package [`nzilbb.vowels`](https://github.com/JoshuaWilsonBlack/nzilbb_vowels).
These functions are designed to apply to any vowel data to which PCA might be
applied. The code in `nzilbb.vowels` is designed to generalise the methods used
in Brand et al. (2021) and Wilson Black et al. (under review). The code for 
the various functions can be found at the GitHub link above and, if desired,
can be directly copied from there if one wishes to modify it and avoid using
the `nzilbb.vowels` package directly.^[
In the course of putting together `nzilbb.vowels`, we have attempted to stick
as closely as possible to what the code would look like outside of a package.
The main difference is the addition of `.data`. This is a way of making it 
explicit that variable names come from a particular data frame rather than the
general environment. They can simply be removed if the code is take outside of 
the package, but it won't do any harm to keep them in.
] If you use code from the package for this kind of analysis, please
cite this paper and Brand et al. (2021).

**NB:** Running this RMarkdown will require around 3GB of RAM.

We will use the following packages:
``` {r, message=FALSE}
# We're working in the tidyverse 'dialect'.
library(tidyverse)

# Used for plots.
library(ggrepel)
library(scales)
library(glue)

# Factoextra used for PCA visualisation
library(factoextra)

# nzilbb.vowels used to share the data and some useful functions.
# Install via github using following commented line. If this does not work,
# ensure that you have the package `remotes` installed.
# remotes::install_github('https://github.com/JoshuaWilsonBlack/nzilbb_vowels')
library(nzilbb.vowels)

# We will fit GAMM models as part of our preprocessing example
library(mgcv)
library(itsadug)

# Plotly for 3D visualisation
library(plotly)

# File path management
library(here)
```

The following is the details of the R installataion and environment used to knit
this markdown:
``` {r}
sessionInfo()
```

# Data

Before turning to PCA, it's worth explaining our data a little further.
The data we are working with come from the Origins of New Zealand English
project (ONZE) (Gordon et al. 2004, 2007). The ONZE corpus provides the longest
record of New Zealand English (NZE) speech recordings in existence, with
speakers born over a 137-year time period (1851-1988). The speech samples
themselves were collected during oral history interviews and sociolinguistic
interviews. The combined corpora contain speech data from over 600 different
speakers from different social backgrounds, ethnic groups and geographical
locations in New Zealand’s South Island. This paper takes processed data from
Brand et al. 2021, which, after filtering, includes 481 speakers (225 female,
256 male), born between 1864 and 1982).

There is extensive evidence to suggest that some of the
vowels of NZE may well show patterns of co-variation over time. The three short
front vowels <span style="font-variant: small-caps;">trap</span>,
<span style="font-variant: small-caps;">dress</span> and 
<span style="font-variant: small-caps;">kit</span> have all changed their place of articulation in
NZE, such that there are now two front vowels which have both raised and fronted
(<span style="font-variant: small-caps;">trap</span> and <span style="font-variant: small-caps;">dress</span>) and one centralized vowel (<span style="font-variant: small-caps;">kit</span>) (Langstrof, 2006). This series
of vowel changes is an example of a push chain (Hay et al., 2015), and is one
place where we might expect to find systematic covariation.

Formant values for the monophthongs were automatically extracted by Brand et al.
using Praat, resulting in comparable measurements across a coherent set of
variables. The initial data set of raw formant measurements of the
12 NZE monophthongs contained over 2 million tokens. This was only possible
because the corpus is in the LaBB-CAT environment (Fromont & Hay 2008; 2012).

Brand et al. then apply an extensive data filtering protocol at the token level
in which a series of conditions were placed on the removal of individual vowel
tokens from the data set (see Brand et al. 2021, supplementary material). FOOT
was removed as its token count was insufficient for our pre-PCA modelling step
(see below) and SCHWA was removed along with all unstressed tokens. This left 10
monophthongs.

Brand et al.'s filtering procedure ran as follows:

1. Remove speakers with missing gender or year of birth information.
2. Remove transcripts that are word lists (interview speech only).
3. Remove tokens where Praat failed to extract F1 and F2.
4. Remove tokens with F1 > 1000 Hz (likely errors).
5. Remove vowel tokens with durations <= 0.01s or >= 3s (likely errors).
6. Remove tokens with phonological length > 25 (likely errors).
7. Remove tokens with a hesitation (affects pronunciation of vowels).
8. Remove tokens where the word is not transcribed.
9. Remove stopwords.^[
Stopword list: 'a', 'ah', 'ahh', 'am', "an'", 'an', 'and', 'are', "aren't", 'as', 'at', 'aw', 'because', 'but', 'could', 'do', "don't", 'eh', 'for', 'from', 'gonna', 'had', 'has', 'have', 'he', "he's", 'her', 'high', 'him', 'huh', 'I', "I'll", "I'm", "I've", "I'd", 'in', 'into', 'is', 'it', "it's", 'its', 'just', 'mean', 'my', 'nah', 'not', 'of', 'oh', 'on', 'or', 'our', 'says', 'she', "she's", 'should', 'so', 'than', 'that', "that's", 'the', 'them', 'there', "there's", 'they', 'this', 'to', 'uh', 'um', 'up', 'was', "wasn't", 'we', 'were', 'what', 'when', 'which', 'who', 'with', 'would', 'yeah', 'you', "you've".
]
10. Remove unstressed vowel tokens.
11. Outlier removal (+- 2.5 SD).
12. Removal of <span style="font-variant: small-caps;">foot</span> tokens (insufficient data).
13. Removal of speakers for whom all vowels heavily overlap (likely errors).
14. Removal of speakers who do not have 5 tokens for each remaining vowel.
15. Removal of tokens with 'l' or 'r' as following segment (standard for NZE).
16. Removal of speakers at the very ends of the year of birth distribution.

We do not go into detail about these steps here as there is nothing special 
about them _in connection with PCA_. They are part of the tool kit of work with
large audio corpora and of statistical modelling in general.^[
Interested readers are directed to the [filtering supplementary from Brand et al. 2021](https://nzilbb.github.io/Covariation_monophthongs_NZE/Covariation_monophthongs_filtering_anon.html),
for the implementation of these steps and more detail concerning their rationale.
]

Figure \@ref(fig:speaker-counts) depicts the count of speakers in the filtered
data by year of birth and gender. (This plot is **Figure 1** of the main 
paper.)
``` {r speaker-counts, fig.cap = "ONZE speakers by year of birth and gender."}
figure_1 <- onze_vowels_full %>%
  group_by(speaker) %>%
  summarise(
    yob = first(yob),
    gender = first(gender)
  ) %>%
  ggplot(
    aes(
      x = yob,
      fill = gender
    )
  ) +
  geom_histogram(bins = 50) +
  labs(
    title = "Speakers by Year of Birth and Gender",
    x = "Year of birth",
    y = "Count of speakers",
    fill = "Gender"
  ) +
  scale_fill_manual(values = c("#f1a340", "#998ec3"))

ggsave(
  here('plots', 'figure_1.png'), 
  figure_1, 
  dpi = 300, 
  width = 160, 
  height = 100,
  units = "mm"
)
figure_1
```

# PCA explainer

_This section provides the initial example to explain PCA in the paper._

We begin by extracting the raw F1 values of <span style="font-variant:
small-caps;">dress</span>, <span style="font-variant: small-caps;">kits</span>,
and <span style="font-variant: small-caps;">trap</span> from a pre-generated
random sample of speakers from the ONZE corpus, with 50 born before 1920 and
50 born after 1920. 

The following code loads the data for this subset of 100 speakers from the
`nzilbb.vowels` package (which includes the dataset `onze_vowels`). It
selects the vowels which we will consider and the useful columns (to see
what the `onze_vowels` dataset looks like before we remove information just
type `onze_vowels` into the console and press enter).^[This assumes that you
are using RStudio to view this document.]
The code block then calculates the mean first formant value for each speaker.
``` {r}
# Load ONZE data for 100 speakers, take only DRESS KIT and TRAP.
onze_sub <- onze_vowels %>%
  filter(
    vowel %in% c("DRESS", "KIT", "TRAP") 
  ) %>%
  select(
    speaker, vowel, F1_50, yob, gender
  ) %>%
  # Take means of first formant data
  group_by(
    speaker, vowel
  ) %>%
  summarise(
    F1_50 = mean(F1_50),
    yob = first(yob),
    gender = first(gender)
  ) %>%
  ungroup()

onze_sub
```

The output above shows that there is a row for each vowel and speaker. That is,
there are three rows for each speaker. However, PCA requires a column for each
variable we are interested in. So we pivot our data so there is a column for
each of the three included vowels.
``` {r}
onze_to_pca <- onze_sub %>%
  pivot_wider(
    names_from = vowel,
    values_from = F1_50
  )
```

At this point we can plot the data, using colour to represent
the third dimension. That is, <span style="font-variant: small-caps;">dress</span>
and <span style="font-variant: small-caps;">trap</span> are represented by
the $x$ and $y$ axes, respectively, while <span style="font-variant: small-caps;">kit</span>
is represented by shade of blue, with darker blues indicating lower values and
lighter blues representing higher values.
``` {r toy-2d, fig.cap = 'Midpoint first formant values for DRESS, KIT and TRAP.'}
three_vowel_plot <- onze_to_pca %>%
  ggplot(
    aes(
      x = DRESS,
      y = TRAP,
      colour = KIT
    )
  ) +
  geom_point() +
  labs(
    title = "First Formant Values",
    x = "DRESS (Hz)",
    y = "TRAP (Hz)",
    colour = "KIT (Hz)"
  )

three_vowel_plot
```

Figure \@ref(fig:toy-2d) is a representation of a 3D cloud of
points. As explained in the paper, PCA can be thought of as placing a point at
the centre of the cloud, and then drawing lines through the cloud, passing
through the central point, which successively capture the largest possible
amount of variance in the cloud with each line at right angles to the previous
line. The lines are what we call principal components (PCs). If you find it hard
to interpret the use of colour in this plot, scroll down for an interactive 
3D scatter plot of the same data which does not require the use of colour.

To show this visually, we will run a PCA analysis using R and plot the results.
We won't pause over the details of the code here as it will be explained in 
more detail in the following section.

We run the PCA analysis using the base `prcomp` function. In this case,
we're not scaling the data, because I want to show what PCA looks like in the
original cloud of points. **NB:** in almost all situations the `scale` argument
to `prcomp` should be set to `TRUE`.^[The reason we do not scale for this example
is that we are trying to visualise PCA in a cloud of points which is as close as possible
to the raw data. Introduction of extra processing steps hinders this.]

``` {r}
# Running PCA
onze_pca <- prcomp(
  # Only select the numeric columns which you want to feed into the PCA.
  onze_to_pca %>% select(DRESS, KIT, TRAP),
  scale=FALSE
)
```

We then extract the required information from the PCA analysis to plot the 
results. Again, we will discuss how to extract this information in more detail
in the following section.
``` {r}
# Load scores for individuals (contained in onze_pca$x)
onze_to_pca <- onze_to_pca %>%
  mutate(
    PC1 = onze_pca$x[, 1],
    PC2 = onze_pca$x[, 2]
  )

# Extract centre of the point cloud.
pca_centre <- onze_pca$center

centre_data <- tibble(
  "DRESS" = pca_centre[[1]],
  "KIT" = pca_centre[[2]],
  "TRAP" = pca_centre[[3]]
)

# Extract loadings.
pca_loadings <- onze_pca$rotation

# Use the loadings and centre to find where each point sits along PC1 and PC2
# when represented in the original 3D space.
onze_to_pca <- onze_to_pca %>%
  mutate(
    PC1_DRESS = (PC1 * pca_loadings[1, 1]) + pca_centre[[1]],
    PC1_KIT = (PC1 * pca_loadings[2, 1]) + pca_centre[[2]],
    PC1_TRAP = (PC1 * pca_loadings[3, 1]) + pca_centre[[3]],
    PC2_DRESS = (PC2 * pca_loadings[1, 2]) + pca_centre[[1]],
    PC2_KIT = (PC2 * pca_loadings[2, 2]) + pca_centre[[2]],
    PC2_TRAP = (PC2 * pca_loadings[3, 2]) + pca_centre[[3]],
  )
```

We plot the centre of the point cloud:
``` {r toy-centre, fig.cap='Midpoint first formant values for DRESS, KIT and TRAP with centre of point cloud.'}
centre_plot <- onze_to_pca %>%
  ggplot(
    aes(
      x = DRESS,
      y = TRAP,
      colour = KIT
    )
  ) +
  geom_point() +
  geom_point(data = centre_data, size = 10, shape = 4, stroke = 2) +
  labs(
    title = "Centre of Point Cloud",
    x = "DRESS (Hz)",
    y = "TRAP (Hz)",
    colour = "KIT (Hz)",
    caption = "The centre of the point cloud is indicated by a cross."
  )

centre_plot
```

Figure \@ref(fig:toy-centre) shows an 'x' where the middle of the point cloud
is. This is clear to see for the <span style="font-variant: small-caps;">dress</span>
and <span style="font-variant: small-caps;">trap</span> axes. Take a moment to 
convince yourself that the colour sits in the middle of the range of colours of
the points for the <span style="font-variant: small-caps;">kit</span> variable.
Don't worry if you can't discern fine differences in the shade of blue, for the
purposes of this illustration all that is required is a sense of 'high', 'mid',
and 'low' values.

We now add the first and second PCs, using line type to visually distinguish
them:
``` {r toy-pcs, fig.cap = 'Midpoint first formant values for DRESS, KIT and TRAP with centre of point cloud and first two PCs.'}
PC1_projections <-  onze_to_pca %>%
  select(
    speaker, PC1_DRESS, PC1_TRAP, PC1_KIT
  ) %>%
  rename(
    DRESS = PC1_DRESS,
    TRAP = PC1_TRAP,
    KIT = PC1_KIT
  )
  
PC2_projections <-  onze_to_pca %>%
  select(
    speaker, PC2_DRESS, PC2_TRAP, PC2_KIT
  ) %>%
  rename(
    DRESS = PC2_DRESS,
    TRAP = PC2_TRAP,
    KIT = PC2_KIT
  )

PC_projections <- bind_rows(
  "PC1" = PC1_projections,
  "PC2" = PC2_projections,
  .id = "PC"
)

pc_plot <- centre_plot +
  geom_line(
    aes(
      x = DRESS,
      y = TRAP,
      colour = KIT,
      size = PC,
      group = PC
    ),
    data = PC_projections
  ) +
  scale_size_manual(
    values = c(
      "PC1" = 2,
      "PC2" = 1
    )
  ) +
  labs(
    title = "First Two PCs",
    caption = NULL
  )

pc_plot
```

Figure \@ref(fig:toy-pcs) might initially be a bit hard to interpret. We see
that PC1 goes from low values of <span style="font-variant: small-caps;">dress, trap</span>, 
and <span style="font-variant: small-caps;">kit</span>. This is easy to see for
<span style="font-variant: small-caps;">dress</span> and <span style="font-variant: small-caps;">trap</span>.
For <span style="font-variant: small-caps;">kit</span>, notice that the thicker
line is dark blue at low values of <span style="font-variant: small-caps;">dress</span>
and <span style="font-variant: small-caps;">trap</span> and light blue at high
values of <span style="font-variant: small-caps;">dress</span> and <span style="font-variant: small-caps;">trap</span>.
Dark blue indicates low values of <span style="font-variant: small-caps;">kit</span>,
while light blue represents high values.

The second PC, however, puts <span style="font-variant: small-caps;">kit</span>
against <span style="font-variant: small-caps;">dress</span> and
<span style="font-variant: small-caps;">trap</span>. This is easiest to see
by noting that the light to dark blue pattern is the opposite of the pattern
for PC1.

The PCs represent new variables against which we can plot our data. We can
reduce this 3D data to two dimensions by plotting the data with respect to PC1
and PC2.
``` {r toy-pc1-pc2, fig.cap="ONZE data for DRESS, TRAP, and KIT, plotted against PC1 and PC2."}
two_dimm <- onze_to_pca %>%
  ggplot(
    aes(
      x = PC1,
      y = PC2
    )
  ) + 
  geom_point() +
  labs(
    title = "2D Representation",
  )

two_dimm
```

Sometimes, plots like \@ref(fig:toy-pc1-pc2) allow for individuals to be
visually clustered into distinct groups. This is the basis of use of PCA as a
prepatory step for classification methods. There are no obvious clusters in the
PC1 and PC2 plot for this this collection of individuals. However, even if there
is no clear grouping, the plot can still be made useful with the addition of
supplementary information (namely, information not fed to the PCA analysis), to
aid in the interpretation of two PCs. For instance:

``` {r toy-pc1-pc2-supp, fig.cap="ONZE data for DRESS, TRAP, and KIT, plotted against PC1 and PC2 with speaker gender and year of birth added as supplementary variables."}
two_dim_supps <- onze_to_pca %>%
  ggplot(
    aes(
      x = PC1,
      y = PC2,
      shape = gender,
      colour = yob
    )
  ) + 
  geom_point() +
  labs(
    title = "PCA Representation of DRESS, TRAP, and KIT F1 Data",
    subtitle = "Supplementary variables: gender and year of birth."
  )

two_dim_supps
```

We will use the information in \@ref(fig:toy-pc1-pc2-supp), in a moment. But
first, to interpret figures \@ref(fig:toy-pc1-pc2) and
\@ref(fig:toy-pc1-pc2-supp), we need to know what the numbers along the PC axes
mean with respect to the original variables. The way to do this is to look at
the 'loadings' of the original variables for each PC. This information is
extracted from the results of `prcomp` as follows:^[This information is **Table
1** in the paper.]

``` {r}
onze_pca$rotation
```

The values for PC1, for each original variable are negative. These indicate
that increases in PC1 correspond to across-the-board decreases in all of 
the original variables. If we look at the column for PC2, we see negative
values for <span style="font-variant: small-caps;">dress</span> and 
<span style="font-variant: small-caps;">trap</span>, but not for 
<span style="font-variant: small-caps;">kit</span>. This means that higher
values of PC2 correspond to lower values for 
<span style="font-variant: small-caps;">dress</span> and 
<span style="font-variant: small-caps;">trap</span>, but higher values
for <span style="font-variant: small-caps;">kit</span>.

\@ref(fig:toy-pc1-pc2-supp) and \@ref(fig:toy-pcs), along with the loadings
we have just considered, give us information which 
allows us to interpret the PCs. Interpretation of PCs is, as the paper explains,
vital for the use of PCA to explore vowel covariation.

If high values of PC1 correspond to low first formant values for all vowels,
then it stands to reason that PC1 is capturing something like vocal tract
length. Speaker gender is an unreliable proxy for this, but a proxy nontheless.
We see the preponderance of male speakers are at high PC1 levels --- i.e., the
majority of male speakers are on the high side of PC1.

PC2, on the other hand, pits <span style="font-variant: small-caps;">dress</span>
and <span style="font-variant: small-caps;">trap</span> against 
<span style="font-variant: small-caps;">kit</span>. It is well known in New
Zealand English, that <span style="font-variant: small-caps;">trap</span>
and <span style="font-variant: small-caps;">dress</span> has risen while
<span style="font-variant: small-caps;">kit</span> has lowered.^[
Keeping in mind that 'rising' is, in fact, a _lowering_ of frequency and
vice versa.
]
PC2 seems to capture this variation. This is confirmed in 
\@ref(fig:toy-pc1-pc2-supp), where we see that the speakers year of birth closely
corresponds to their PC2 score. The later you are born, the higher your PC2 
score.

In a full research project, it is often useful to provide further evidence of a
connection between PC scores and this or that additional variable. Here, any
of the usual tools of statistical modelling can be used. For instance, we might
fit a linear model of PC1 score by year of birth to confirm the relationship
we set out visually above.

So, we have seen how we go from 3D data from three distinct vowels to a 2D
representation of the data in which the new dimensions are _interpretable_. This
is, in small form, the benefit of PCA.

For the purposes of the paper, we combine four the the above plots as a single
plot using the `patchwork` package. This is **Figure 2** in the paper.
``` {r, fig.dim = c(10, 10), fig.cap = "Figure 2 for paper."}
figure_2 <- (three_vowel_plot + centre_plot) /
  (pc_plot + fviz_screeplot(onze_pca, barcolor = "darkgreen", barfill = "darkgreen")) +
  plot_layout(guides = "collect") +
  plot_annotation(
    title = "PCA Applied to NZE Short Front Vowel First Formants",
    tag_levels = "A",
    theme = theme(plot.title = element_text(size = 20))
  )

ggsave(
  here('plots', 'figure_2.png'),
  figure_2,
  dpi = 500, 
  width = 320, 
  height = 200,
  units = "mm"
)
figure_2
```



**Figure 3**, in the paper, will be the 2D representation of the data with and
without supplementary variables:
``` {r}
figure_3 <- (two_dimm + labs(title = NULL)) + 
  (two_dim_supps + labs(title = NULL, subtitle = NULL)) +
  plot_annotation(
    title = "PC1 and PC2 Representation of Data",
    subtitle = "With and without supplementary variables.",
    tag_levels = "A",
    theme = theme(
      plot.title = element_text(size = 20),
      plot.subtitle = element_text(size = 15)
    )
  )

ggsave(
  here('plots', 'figure_3.png'),
  figure_3,
  dpi = 300, 
  width = 320, 
  height = 150,
  units = "mm"
)
figure_3
```

Finally, we provide the code for a 3D interactive plot which enables one to
visualise the steps carried out by PCA geometrically. This is a long code block,
which is hidden by default. To view the code, press the 'Code' button. The
resulting figure enables one to move around the point cloud, add the central
point, and the first two PCs (use the box at the top right to show or hide
these elements from the plot).
``` {r 3d-fig, class.source='fold-hide'}
# Visualisation

fig_3d <- plot_ly(
  data = onze_to_pca,
  x = ~DRESS,
  y = ~KIT,
  z = ~TRAP,
  type = "scatter3d", # What kind of plot.
  mode = "markers", # What kind of marking (something like geom in ggplot)
  name = "Speakers", # This label will go in the legend.
  marker = list(
    size = 2,
    opacity = 1,
    color = "gray",
    lines = list(
      color = 'black',
      width = 1
    ),
    showlegend = FALSE
  ),
  hoverinfo = 'text',
  text = ~paste0(
    "DRESS: ", round(DRESS), " (Hz)<br>", 
    "KIT: ", round(KIT), " (Hz)<br>", 
    "TRAP: ", round(TRAP), " (Hz)" 
  ),
  projection = list(
    x = list(
      show = TRUE,
      scale = 0.2
    ),
    y = list(
      show = TRUE,
      scale = 0.2
    ),
    z = list(
      show = TRUE,
      scale = 0.2
    )
  )
)

# Add title and axis labels
fig_3d <- fig_3d %>%
  layout(
    scene = list(
      xaxis = list(
        title = "DRESS F1 (Hz)"
      ),
      yaxis = list(
        title = "KIT F1 (Hz)"
      ),
      zaxis = list(
        title = "TRAP F1 (Hz)"
      )
    ),
    plot_bgcolor = 'rgba(0, 0, 0, 0)',
    paper_bgcolor = 'rgba(0, 0, 0, 0)'
  )

fig_3d <- fig_3d %>%
  # Add centre mark
  add_trace(
    x = ~DRESS,
    y = ~KIT,
    z = ~TRAP,
    data = centre_data,
    type = 'scatter3d',
    mode = 'markers',
    marker = list(
      size = 6,
      opacity = 1,
      color = 'red',
      symbol = "x"
      ),
    name = "Centre",
    inherit = FALSE,
    visible = "legendonly",
    hoverinfo = 'text',
    text = ~paste0(
      "DRESS: ", round(DRESS), " (Hz)<br>", 
      "KIT: ", round(KIT), " (Hz)<br>", 
      "TRAP: ", round(TRAP), " (Hz)" 
    )
  ) %>%
  # Add PC1 line
  add_trace(
    data = onze_to_pca,
    x = ~PC1_DRESS,
    y = ~PC1_KIT,
    z = ~PC1_TRAP,
    type = 'scatter3d',
    mode = 'lines',
    name = 'PC1',
    line = list(
      color = "#D55E00",
      width = 6
    ),
    inherit = FALSE,
    visible = "legendonly",
    hoverinfo = 'text',
    text = ~paste0(
      "DRESS: ", round(PC1_DRESS), " (Hz)<br>", 
      "KIT: ", round(PC1_KIT), " (Hz)<br>", 
      "TRAP: ", round(PC1_TRAP), " (Hz)" 
    )
  ) %>%
  # Add PC2 line.
  add_trace(
    data = onze_to_pca,
    x = ~PC2_DRESS,
    y = ~PC2_KIT,
    z = ~PC2_TRAP,
    type = 'scatter3d',
    mode = 'lines',
    name = 'PC2',
    line = list(
      color = "#0072B2",
      width = 4
    ),
    inherit = FALSE,
    visible = "legendonly",
    hoverinfo = 'text',
    text = ~paste0(
      "DRESS: ", round(PC2_DRESS), " (Hz)<br>", 
      "KIT: ", round(PC2_KIT), " (Hz)<br>", 
      "TRAP: ", round(PC2_TRAP), " (Hz)" 
    )
  )

fig_3d
```

# Case Study: Brand et al. 2021

The following code repeats, with some modifications, the analysis of Brand et
al. (2021).^[
All data and analysis from Brand et al. (2021) can be found online either at the following GitHub repository: https://github.com/nzilbb/Covariation_monophthongs_NZE or at The Open Science Framework at: https://osf.io/q4j29/.
Individual files can be accessed at https://nzilbb.github.io/Covariation_monophthongs_NZE/. We note, 
in particular, the online version of the analysis supplementary https://nzilbb.github.io/Covariation_monophthongs_NZE/Covariation_monophthongs_analysis.html.
]

The analysis is divided into the following steps:

1. data processing,
2. application of PCA (including PCA tests), and
3. interpretation of PCs.

We use all 481 speakers for this analysis, rather than the sample of 100 used in
our explanation of PCA above.

## Data Processing

_This section corresponds to the section 'Data processing and research_
_questions' in the paper._

For PCA to tell us anything interesting, we need to put in the right kind of 
data. In the case of across-speaker changes in vowel shapes we are not 
interested in changes due to difference in vocal tract length. For this reason,
we apply a normalisation method to the data. The method is a modification of the
Lobanov normalisation, changed to allow for differences in the proportion of
tokens of each vowel. This means we can use Lobanov normalisation for
naturalistic, rather than controlled laboratory recordings.

We then take a step beyond normalisation, by fitting models to capture and then
exclude known  sources of variation in vowel data. PCA should then help us to
find vowel covariation which is not explained by known sources of variation.

We'll take each step in turn.

### Normalisation

We use the Lobanov 2.0 normalisation, as developed in Brand et al. (2021).
We give the formula here:
$$
\begin{equation}
F_{lobanov_i} = \frac{(F_{raw_i}-\mu_{raw_i})}{\sigma_{raw_i}}
\end{equation}
$$

Lobanov 2.0 is implemented in the R package `nzilbb.vowels`. The 
function `lobanov_2` assumes that the first column contains
speaker identifiers, the second column contains vowel identifiers, and the
third and fourth columns contain, respectively, F1 and F2 data in Hertz.^[
This matched the structure used in the `vowels` package.
]
We the apply function as follows:
``` {r, message = FALSE}
onze_vowels <- onze_vowels_full %>%
  # Add F1_lob2 and F2_lob2 columns
  lobanov_2() %>%
  # Remove F1_50 and F2_50 columns
  select(-(F1_50:F2_50))
```

### Modelling

We want to fit models which capture the variation in the data which we are _not_
interested in. We are looking for covariation which is _not_ 
explained by changes in our predictor variables.
Here we fit a distinct GAMM model to each vowel and formant pair
which captures differences explicable by gender, by speech rate, and by the
speaker's year of birth. We also fit a random intercept for each speaker and for
each word. In order to speed up this process, we remove any
words which have less than five tokens. This is not done by Brand et al. (2021).

The models are here fit using the methods provided by `purrr`, the tidyverse
package for functional programming.^[A blog post setting out this method step
by step can be found 
[here](https://joshua.wilsonblack.nz/post/visualising-vowel-space-gamms/).]
Each model if fit using the `bam` function from `mgcv`, using the following
formula: `formant_value ~ gender + s(yob, by=gender) + s(speech_rate) + s(speaker, bs="re") + s(word, bs = "re")`
This is done separately for each vowel and formant. The smooths are fit using the
default 10 knots. There is a parametric term for `gender`, and a distinct smooth
is fit for each value of `gender`. This means that the two genders are fit 
separately. Brand et al. (2021) represent the smooths using 'difference smooths'
for `gender` rather than fitting two independent smooths. This introduces an
unnecessary complication for our purposes here and will not affect the random
intercepts which we intend to extract from these models at a later stage.^[
Recall that this is not an explainer on GAMMs. For that, see the bibliography to 
the paper.
]

``` {r, cache = TRUE}
# Nest the data so that there is a single row for each model.
onze_models <- onze_vowels %>%
  pivot_longer(
    cols = matches("F[1-2]"), # Select columns with formant data (assumes 
    # there are only two of these)
    names_to = "formant_type", # Name column to distinguish F1 & F2
    values_to = "formant_value" # Name column for formant values.
  ) %>%
  group_by(vowel, formant_type) %>%
  nest() %>%
  ungroup()

# For each dataset, remove words with less than five appearances.
onze_models <- onze_models %>%
  mutate(
    data = map(
      data,
      ~ .x %>%
        group_by(word) %>%
        mutate(
          n_occurrences = n()
        ) %>%
        filter(
          n_occurrences >= 5
        )
    )
  )

# Fit GAMMS
onze_models <- onze_models %>%
  # For each vowel-formant pair, fit an lmer model.
  mutate(
    model = map( # map takes...
      data, # ... each entry in a column, and ...
      ~ bam( # applies a function to it.
        formant_value ~ gender + s(yob, by=gender)
        + s(speech_rate) + s(speaker, bs="re") + s(word, bs = "re"),
        discrete = TRUE, nthreads = 16,
        data = .x # where .x is a pronoun referring to the column entry.
      )
    )
  )
```

We now have models for each vowel and formant pair. We can have a quick look 
at the first of these models to illustrate what they look like.  
This model is for the <span style="font-variant: small-caps;">`r onze_models$vowel[[1]]`</span> vowel's 
`r onze_models$formant_type[[1]]` values.
``` {r}
summary(onze_models$model[[1]])
```
The GAMM output indicates we have a significant smooth over year of birth for
both male and female speakers. We are also capturing a lot
of variance with our smooths for each speaker and for each word.

Let's see what the smooths look over year of birth (using the `plot_smooth` 
function from `itsadug`:
``` {r}
plot_smooth(onze_models$model[[1]], view = "yob", plot_all = "gender", rug = T)
```

According to this model, at the start of our time period, men and women are
distinguishable in the height of their 
<span style="font-variant: small-caps;">thought</span> vowel at the start and
end of our period, but are not distinguished in a period roughly from 1920 to
1940. The rug at the bottom of the above plot indicates
when our actual speakers were born.

We now have 20 models. But we want to do PCA with _speakers_ as rows. For this,
we need a matrix with a row for each speaker. What are the values in the
matrix going to be?

One option would be to get each model's _predictions_ for each speaker. This 
would give us the variation _explained by_ the model predictors. But this is not
what we want for our project.^[It _is_ a reasonable option for sociolinguistic
investigation of vowels in which we want to fix our attention on a certain 
kind of variation, decided upon in advance. This approach would be an extension of 
'canonical correspondence analysis' or (CCA). For more about CCA in sociolinguistics
see [Walker et al. (2022)](https://www.taylorfrancis.com/chapters/edit/10.4324/9781003134558-7/lect-coherence-phonetic-grammatical-variation-james-walker-michol-hoffman-miriam-meyerhoff).]
We, instead, want the variation which is _not_ captured by these predictors.
One way to get this is to extract the random intercepts which the model fits
to each speaker. These are the model's way of capturing the left over variance
introduced by each speaker, assuming that speakers come from a natural population
of possible speakers.^[For an account of how this works with linear mixed
models, see [Drager and Hay (2012)](https://www.proquest.com/docview/1008896594?parentSessionId=%2Fjqcqigo8JxO%2F4M%2BgKc3WbsiKKRPZTCzCaCOIm810zE%3D).]

We pick out these random effects using the function `get_random()`
from the package `itsadug`. After applying
this function to each model, we can 'unnest' to get a single matrix with the
random intercepts for each speaker, for each vowel, and for each formant
type.
``` {r}
onze_models <- onze_models %>%
  mutate(
# For each model, extract random effects and turn them into a data frame
    random_intercepts = map(
      model,
      ~ get_random(.x)$`s(speaker)` %>% as_tibble(rownames = "speaker") %>%
        rename(intercept = value)
    )
  )

# Select the vowel, formant type, and random intercepts columns and 'unnest'
onze_ints <- onze_models %>%
  select(vowel, formant_type, random_intercepts) %>%
  unnest(random_intercepts)

onze_ints
```

This isn't quite what we want, yet. The data is too 'long'. It needs to be 
wider in the sense that we need a column for each vowel and formant type pair (
i.e., we need a column for <span style="font-variant: small-caps;">thought</span>
F1 and another for <span style="font-variant: small-caps;">thought</span> F2
etc.
).

We widen the data as follows:

``` {r}
onze_ints <- onze_ints %>%
  # Create a column to store our eventual column names
  mutate(
    # Combine the 'vowel' and 'formant_type' columns as a string.
    vowel_formant = str_c(vowel, '_', formant_type),
    # Remove '_lob2' for cleaner column names
    vowel_formant = str_replace(vowel_formant, '_lob2', '')
  ) %>%
  # Remove old 'vowel' and 'formant_type' columns
  select(-c(vowel, formant_type)) %>%
  # Make data 'wider', by...
  pivot_wider(
    names_from = vowel_formant, # naming columns using 'vowel_formant'...
    values_from = intercept # and values from intercept column
  )

onze_ints
```


## Application of PCA

_This section corresponds to the section 'PCA' in the paper._

We will begin by applying PCA and looking at its output before turning, more 
seriously, to the answers to the three main questions we need to ask in 
any PCA analysis:

1. Is PCA appropriate for this data.
2. How many PCs should we consider.
3. Which variables should we use to interpret the PCs.

Our data is in the required form for PCA: that is, it has rows for each
observation (in our case, a single row for each speaker), and has values in
each column (there are no missing values in our data).

The base R function `prcomp` takes a matrix of numerical values. This means
we need to remove our speaker identifiers from the data when we pass it to 
`prcomp`.

``` {r}
onze_pca <- onze_ints %>%
  # Remove the speaker column as this is not wanted by PCA.
  select(-speaker) %>%
  # We scale the variables (more on this in a moment)
  prcomp(scale = TRUE)
```

It is usually recommended to scale the data (which we did not do in the 3D
example above). If we don't scale the data, then those variables which have the
largest variance will dominate.^[
This is a bigger problem in applications where
the data comes from variables which are one quite different scales. Say, for
instance, that we added speaker age as a variable and didn't scale it. If we had
even a small range of speaker ages, these would quickly dominate over Lobanov
2.0 normalised vowel space data (where, e.g. the <span style="font-variant:
small-caps;">thought</span> F1 values runs from  
`r min(onze_models$data[[1]]$formant_value)` to 
`r max(onze_models$data[[1]]$formant_value)`).
]
If you don't have a positive reason to leave the data unscaled, then ensure
you add `scale=TRUE`. **NB:** variable scaling is not the default behaviour for
`prcomp`. This is for backwards compatibility reasons.

Often the first question about a PCA analysis is how much variance each of the
PCs explains. PCA attempts to find new variables, the PCs, which explain, 
successively, as much of the variance in the original data as possible. For
that reason, we want our first few PCs to explain the majority of the variation
in the data. We visualise the variance explained with a scree plot using the
function `fviz_screeplot` from `factoextra`.
``` {r}
fviz_screeplot(onze_pca)
```

We have a PCs along the x axis, and the percentage of variance they explain
along the y-axis. This plot only shows the first 10 PCs.^[There are 20 PCs in
total. PCA will produce as many PCs as there are variables or observations
(whichever number is smaller).] Our first PC explains around 17% of the variance
in the data. This is more than we would expect if each variable were covering an
equal amount of the variance in the data. If the variance were evenly split
across the variables, we would expect them to all account for (1/20 = `r 1/20 *
100`%) of the data. In fact, we see that the first few PCs all explain more than
5% of the variance, while all the PCs after PC7 explain less than 5% of the
variance in the data.

The amount of variance explained by each PC is captured in the PCA object (i.e.,
the output of `prcomp`) by `$sdev`. These are the square roots of the
eigenvalues of the correlation matrix of the data. That is, we are picking out
structure in the correlations which are present in the data. We turn these
eigenvalues into percentages of variance explained by squaring the values and
dividing each by the sum of the squared variables (and multiplying by 10).
In R:
``` {r}
onze_pca$sdev^2 / sum(onze_pca$sdev^2) * 100
```

These values in the output above correspond to the percentages given in the
scree plot above.

As we saw above, PCA generates 'loadings', which tell us how the original 
variables relate to the PCs and 'scores' which situate each observation
with respect to the PCs. We can access there from the output of `prcomp`
using `$rotation` for the loadings and `$x` for the scores.

``` {r}
onze_pca$rotation
```
The output above gives the loadings for all 20 PCs. Inspection of the first
column reveals that <span style="font-variant: small-caps;">thought</span>
 F2 is given the highest loading for PC1. 

We can look at the first few scores too:
``` {r}
head(onze_pca$x)
```
This tells us where the individuals sit within the space defined by
the principal components. So, individual 1 (the first row), has positive
values for the first five PCs, but the magnitude of their PC scores is quite
small compared to some other speakers (e.g. compare the PC1 value for the sixth
row and the PC2 value for the speaker in the second row).

For the purpose of illustration, let's look at a popular method for depicting
loadings: the variable plot. This allows us to plot the original variables 
within the space defined by any two of the PCs. The natural choice is to use
the first two principle components. For this we use the function `fviz_pca_var`
from `factoextra`.
``` {r variable-plot, fig.cap = "Variable plot of PC1 and PC2."}
fviz_pca_var(
  onze_pca,
  # top 10 variables by contribution to the PCs plotted
  select.var = list(contrib = 15),
  repel = TRUE # attempt to prevent label overlap
)
```

Here we see PC1 along the x axis and PC2 along the y axis. Each label also tells
us what proportion of the variance in the data these PCs explain (the same 
information presented in the scree plot above). If two arrows point in opposite
directions, they are patterned against each other in the space of these PCs,
if they point in the same direction, then they are patterned with one another.
The closer the arrows are to the circle, the better they are represented by 
these two PCs. That is, if an arrow were to touch the circle, then no
further PCs would be required for us to perfectly capture it's position in our
original data set. It is to be expected that we lose some quality of representation
when doing PCA. In fact, this is desired if we think of later PCs are capturing
noise in our original data and earlier PCs as capturing signal.

If we look along the x-axis, we see <span style="font-variant: small-caps;">fleece</span> and
<span style="font-variant: small-caps;">thought</span>
F2 and <span style="font-variant: small-caps;">trap</span> and <span style="font-variant: small-caps;">thought</span>
and against <span style="font-variant: small-caps;">start</span> and <span style="font-variant: small-caps;">fleece</span>
F1. If we look along the y axis, we see <span style="font-variant: small-caps;">kit</span>
and <span style="font-variant: small-caps;">strut</span> F2 patterned against
<span style="font-variant: small-caps;">dress, trap</span> and <span style="font-variant: small-caps;">thought</span>
F2. This plot will be returned to later, as we attempt to interpret our PCs in 
the face of instability in their direction.

We can follow up this PC-by-PC approach to interpreting the loadings with 
contribution plots. These can be carried out using the `pca_contrib_plot` 
function from `nzilbb.vowels`.
``` {r pc1-contrib, fig.cap="Contribution plot of PC1."}
pca_contrib_plot(onze_pca, pc_no = 1, cutoff = NULL)
```

This shows the _contribution_ of each variable to the PC, measured as a
percentage. This is an alternative to the raw loading. We indicate whether the
underlying loading is negative or positive by use of either a '+' or '-' and
by colour. Here, we see the vowels we picked out in the variable plot. This plot
makes the relative magnitude of different variables more clear than if 
we used only the variable plot.

We can plot the same plot with a 'cut off' value. This is a method which is 
used sometimes when interpreting PCs.
``` {r pc1-contrib-cut, fig.cap="Contribution plot of PC1 with 50% cutoff."}
pca_contrib_plot(onze_pca, pc_no=1, cutoff=50)
```

Figure \@ref(fig:pc1-contrib-cut) hides those variables which are not required to account for 50% of the
contribution of the original variables to the PC.

We now show the contribution plot for PC2.

``` {r pc2-contrib, fig.cap="Contribution plot of PC1."}
pca_contrib_plot(onze_pca, pc_no = 2, cutoff = NULL)
```

We provide the contribution plot of PC2 in order to enable comparison between
it and PC1 and with the first two PCs as represented in the variable plot 
(Figure \@ref(fig:variable-plot))

Individuals can also be plotted in the space of the first two PCs (or, indeed
any two PCs we select). We can do this using the function `fviz_pca_ind` 
from `factoextra`.
``` {r}
fviz_pca_ind(
  onze_pca,
  repel=TRUE
)
```

It is often useful to understand how the observations in a data set are being
placed in the space defined by a given set of PCs. In this case, we see that
the individual in the 150th row of the data has a very high PC2 score and a 
reasonably low PC1 score. This individual, and the individual in the 152nd row
are coming out as quite extreme in value for PC2.
We will discuss how this information can be used further
below. 

Finally, both variables and individuals can be joined in a 'biplot', 
using `fviz_pca_biplot`. In this case, we restrict both the individuals and the
variables to those which make the biggest contribution to the two PCs. We pick
10 of the original variables, and 30 of the individuals. Only the 10 variables
which most contribute to their PCs and the 30 speakers who take the most extreme
values for them are plotted.
``` {r variable-biplot, fig.cap = "Biplot of PC1 and PC2."}
fviz_pca_biplot(
  onze_pca,
  select.var = list(contrib = 10),
  select.ind = list(contrib = 30),
  repel = TRUE
)
```
Sometimes it can be useful to have the information in the variable plot and 
the individual plot at the same time.

### Is PCA Appropriate?

The first question to ask concerning a dataset which we intend to apply PCA to
is, straightforwardly, "is PCA appropriate?". This is particularly important
given that PCA will _always_ find relationships between variables. It is 
important to be able to determine whether these relationships represent genuine
phenomena or whether they are merely the result of randomness in the data.

We saw above that PCA can be thought of as exploring the structure of
correlations within a dataset. We can thus ask whether there is any such
structure. That is, we can look at the correlations present in our data and
those we would expect if the data lacks an interesting structure in its
correlations (or at least, structure which we can be confident is not random).

There are various ways of comparing the structure of correlations in a data set
compared with randomised data.^[For more options see [(Vieira 2012)](http://www.iaees.org/publications/journals/ces/articles/2012-2(2)/permutation-tests-to-estimate-significances.pdf)
and for implementation in R see [Carmargo 2022](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC8858582/).]
For this question,  will just look at the raw count of significant correlations,
comparing them in our real data and in 1000 permuted versions of the dataset.
Here we use the function `correlation_test` from the `nzilbb.vowels` package.

``` {r, cache = TRUE}
onze_cor_test <- correlation_test(
  onze_ints %>% select(-speaker), 
  n = 1000,
  cor.method = "pearson"
)
```

The basic information about the correlation test can be extracted using
`summary`. Note that you can change the alpha level here.
``` {r}
summary(
  onze_cor_test,
  alpha = 0.05,
  n_cors = 5 # Increase this to see more pairwise correlations.
)
```
The strongest correlation in our data is between <span style="font-variant: small-caps;">strut</span>
and <span style="font-variant: small-caps;">thought</span> F2. There are 129 
significant correlation in the original data, and an average of 9.5 across our
1000 permutations (with a max of 22). This indicates that we have much more
structure in our correlation matrix than we would expect from random data.
We also see strong correlations _across_ different vowels. Only one of the
top five correlations is of an F1 and F2 in a single vowel. Such correlations are,
obviously, not particularly interesting if we are interested in co-variation 
across multiple vowels.

Information about the pairwise correlations which are present in our data can be
useful for interpreting PCs. We will return to this below.

We can plot the correlation test in two ways using `nzilbb.vowels`. First, we
plot the _count_ of significant correlations.
``` {r correlation-check, fig.cap = "Count of significant correlations."}
sig_cor_plot <- plot_correlation_counts(onze_cor_test, half_violin = TRUE)
sig_cor_plot
```

Figure \@ref(fig:correlation-check) presents the count of significant correlations
in the original data as a red dot, and the distribution of counts of significant
correlations as a blue violin.

We can also look at the magnitude of correlations across our variables.
``` {r correlation-check-2, fig.cap = "Magnitude of correlations in original and permuted data."}
mag_cor_plot <- plot_correlation_magnitudes(onze_cor_test)
mag_cor_plot
```
From the above plots, we see that we have non-random structure in our
correlations, both in terms of magnitude and statistical significance. **This**
**gives us good reason to proceed with PCA.** If we could not easily distinguish
our data from the distribution of permuted data sets, we would not be in a
position to apply PCA.

The following plot is **Figure 4** in the paper:
``` {r}
figure_4 <- (sig_cor_plot + labs(title = NULL, caption=NULL)) + 
  (mag_cor_plot + 
     labs(title = NULL, x = "Magnitude of Pearson's r") +
     theme(legend.position = "none")) +
  plot_annotation(
    title = "Permutation Test on By-Speaker Random Intercepts",
    subtitle = "(1000 Iterations)",
    tag_levels = "A",
    theme = theme(
      plot.title = element_text(size = 20),
      plot.subtitle = element_text(size = 15)
    )
  ) +
  plot_layout(guides = "collect")

ggsave(
  here('plots', 'figure_4.png'),
  figure_4,
  dpi = 300, 
  width = 320, 
  height = 150,
  units = "mm"
)
figure_4
```

**Recommendation:** PCA should only be run on vowel data which contains
non-random structure in their pairwise correlations as revealed by permutation
test.

### How Many PCs?

As noted above, when scree plots were introduced, we hope to capture the majority
of variance in our data with the first 'few' PCs. The best method for picking
what counts as a 'few' is not a settled matter. All methods turn on the kind of
information present in a scree plot.

In Brand et al. (2021) adopts two criteria:

1. The final PC we select must explain more variation than would be expected
from the same PC with permuted data. For example, if we are looking at PC2, does
the amount of variance explained by our PC2 exceed that which we might expect
from PC2 in a permuted version of our data set.
2. The final PC we select must explain more than 10% of the variation in the
data.

The use of a cut off for variance explained is quite widespread in the 
literature. For instance, Kaiser 1961 argues that eigenvalues greater than one
should be accepted as they account for more variation than would be explained by
one of the original variables (Kaiser 1961). This would correspond, in our 
dataset, to 5% of variance explained (recall that we have 20 variables).

We believe the method in Brand et al. (2021) remains a defensible one. 
The cut off of 10% can be thought of as a threshold on effect size. There is no
reason not to adopt such a threshold in exploratory work. 
However, in presenting the use of PCA for the
investigation of vowel co-variation is a _general_ method, it is worth offering
a more robust approach which takes account of recent methodological advances 
in the use of PCA in the biological sciences.

In particular, 
permutation and bootstrapping methods have become increasingly popular with the
increased computer power available to researchers using ordinary PCs. Permutation
methods are useful for determining a _null_ distribution. The distribution in
the previous section (Figure \@ref(fig:correlation-check)) is a null
distribution for the count of significant pairwise correlations. As the count
from the original data falls well outside this distribution, we take it that
the correlation structure in our data is _not_ the result of merely random
variation but reflects real structure. Bootstrapping, on the other hand, 
generates new data sets with subsets of the original data and are used to
estimate confidence bands around both the variance explained by each PC and on
each loading.^[Our use of bootstrapping and permutation tests to estimate confidence
bands and null distributions for PCA follows the work of Vieria (2012) and
Carmago (2022), although our implementation in R is distinct from Carmago's.]

The function `pca_test` from `nzilbb.vowels` runs a parallel bootstrap and
permutation test of the sort just described:
``` {r, cache = TRUE}
onze_test <- pca_test(
  onze_ints %>% select(-speaker), 
  n = 1000,
  variance_confint = 0.95,
  loadings_confint = 0.9
)
```

We can directly extract judgements about which PCs are significant and which 
loadings are significant from `pca_test` using the `summary` function. For now,
just look at the information about which PCs are significant:
``` {r}
summary(onze_test)
```

According to the summary above, PCs 1-5 are significant, in the sense that they
explain more variance than attained in 95% of the permuted analyses. This is, 
in practice, equivalent to the first criterion in Brand et al. (2021). 

We can visualise the test results for variance explained using 
`plot_variance_explaind` from `nzilbb.vowels`:
``` {r}
plot_variance_explained(onze_test)
```

Here, we see that PCs 1, 2, 3, and 4 are well clear of the permuted data, while
our value for PC5 is very close to the distribution of permuted values. We also 
see that the confidence interval for PC3 falls either side of 10%. This is a 
worry for using a 10% cut off, as in the second criterion used by Brand et al.
Sometimes this will appear above the threshold and sometimes it won't (this will
be, judging from the plot, a roughly 50/50 proposition).

A more serious issue is revealed by the bootstrap analysis of our data. This was
not clear in the analysis of Brand et al. (2021), but, as we will show below, 
does not affect its conclusions. This is the fact that the confidence bands for
PC1 and PC2 overlap. This is worrying because in situations in which two PCs
explain the same, or nearly the same, variance, they become very unstable. In
the case where they are equal, any two orthogonal lines through the variable
plot of PC1 and PC2 would do equally well for the PCs. This, in turn, leads to 
instability in the loadings for each PC. In fact, some researchers in biology
argue that PCs with overlapping confidence intervals 
_should not be interpreted_ (e.g. [Björklund (2019)](https://onlinelibrary.wiley.com/doi/full/10.1111/evo.13835)).
We believe that this is an excessive response to the problem and will offer a
solution in the following section. However, we agree with Björklund that 
some indication of the error in our estimation of variance explained should
be provided.

This information can be extracted from the `pca_test` object, for
variance explained as a percentage, as follows:
``` {r}
onze_test$variance %>% 
  filter(sig_PC) %>% 
  select(
    PC, variance_explained, low_confint_var, high_confint_var, 
    mean_confint_var, sd_confint_var
  ) %>% 
  # Express as percentage
  mutate(across(.cols = !PC, .fns = ~ .x * 100))
```


**Recommendation:** report on each PC whose values appear as significantly
greater than expected from the null distribution. This could be in supplementary
material, or in the main body text. Also report, either visually or in a table,
the mean, standard deviation and confidence intervals on the variance explained
by each PC.

The following is **Figure 5** in the paper. We use the `pc_max` argument to 
restrict out plot to the first 10 PCs.

``` {r}
figure_5 <- plot_variance_explained(onze_test, pc_max = 10) + 
  labs(caption = NULL)

ggsave(
  here('plots', 'figure_5.png'),
  figure_5,
  dpi = 500, 
  width = 160, 
  height = 100,
  units = "mm"
)
 
figure_5
```

### Interpretation: Which variables?

We now look at interpreting our PCs. We saw in the 3D toy model that PCs can 
have natural explanations. Often, introductions to PCA suggest that the mark of
an interpretable PC is that it has 'high' loadings for a handful of variables.^[
What counts as a 'high' loading depends on the number of variables. The sum of
squares of the loadings is one, so any loading above $\frac{1}{n}$, where n
is the number of variables, is contributing more than would be expected if 
each contributed equally. This is usually a good lower bound for a 'moderate' 
loading.
]
This is just a heuristic though. Some patterns across lots of variables can be
easily interpreted. For instance, if we fed raw hertz values of all of our 
vowels into PCA, we would get something like PC1 in the 3D example, in which
all variables patterned together. This would be just as easy to interpret as a
size factor as it was in the 3D case!^[
**NB**: in this case all of many of the variables would not have 'high' loadings
either. So even the desire to only use high loadings is only a heuristic.
]

In all cases, we will be looking for some pattern which we can explain,
occurring across the loadings. A necessary step for this is to decide which
variables we want to include in our interpretation and which we want to exclude.
There are a few reasons for this:

1. We _exclude_ some variables to avoid reading in to 'noise' in our data.
Variables with loadings which could be explained by random chance are to be 
avoided.
2. We want to discipline our interpretation to avoid including or excluding 
variables to suit a pre-existing bias.
3. We want clear rules which can be applied across multiple data sets or 
research programmes.
4. We want to _include_ all those variables which reliably contribute to the PC.

Brand et al. (2021) adopt a cut off value to interpret each PC (as depicted
above in Figure \@ref(fig:pc1-contrib-cut)). The cut off is applied to the
loadings represented as contribution values (i.e. as percentages). They 
successively take the variables with the highest contribution to the PC until
50% of the contribution of the original variables to the PC is accounted for.
This approach satisfies the reasons just given.

Other options are possible. We present one here which is somewhat more robust
and generalisable.
The 50% cut off value was chosen as a result of trial and error with the 
particular data. In practice, this method results in a very similar collection
of variables to use in interpreting our PCs.

We follow Vieira's (2012) use of   
index loadings. For each loading, for a given PC, the index loading 
is the product of the PC's squared eigenvalue and the squared loading. Index
loadings solve two problems:

1. Index loadings are always positive, whereas loadings can be either positive
or negative. This enables comparison across multiple PCA analyses, where the
sign can swap while the PCs are essentially equivalent.
2. Index loadings take into account the amount of variance explained by a 
given PC. This means that moderate loadings on PCs which explain a lot of 
variance are ranked higher than moderate loadings on PCs which don't explain
much variance, etc.

We use permutation and bootstrapping to generate a 
null distribution and confidence bands for the index loadings for each PC.
Vieira suggests that the standard 0.05 alpha value, and the corresponding use
of 95% confidence intervals, is too conservative when using index loadings. So,
by default, the `pca_test` function uses 90% confidence intervals.

We plot the index loadings for the first PC using the `plot_loadings` function:
``` {r pc1-loadings, fig.cap = "PC1 index loadings with confidence bands and null distribution."}
pc1_loadings_plot_uf <- plot_loadings(onze_test, pc_no=1, filter_boots = FALSE)
pc1_loadings_plot_uf
```

Just as in the plot of variance explained, this plot has both a null distribution
and confidence intervals. The actual values from our PCA analysis are presented
by either a black plus or minus, indicating the sign of the loading. If a black
minus or plus is outside of the interval for the null distribution, then it
is 'significant'. The red bars indicate the 95% intervals for the index loading.

Figure \@ref(fig:pc1-loadings) clearly shows the problem of instability in our 
PCs due to the overlap of the confidence intervals in variance explain for PC1
and PC2. One intuitive way to see this is to imagine that there are two sources
of variation in our data, but that in some subsets of the data one dominates the
other, while in other subsets it is the other way around. In this kind of case,
sometimes PC1 will represents the one source of variation and sometimes it 
will represent another. Given this, we would expect our confidence bands on 
index loadings to include very low values for all our of original variables. 
This is exactly what we find in Figure \@ref(fig:pc1-loadings). Björklund
would here say that this is reason to not apply PCA to this data full stop.

However, we argue that this is too fast. There are genuine relationships of 
covariation in PC1 and PC2 which we want to capture. One way to do this is to 
look at a _subset_ of bootstrap analyses, namely, those in which the variable
with highest median across all bootstrapped analyses takes a value in the top
3/4 of its distribution. This filtering is applied using the `filter_boots`
option (which can take some time, as confidence intervals are recalculated).

``` {r filtered-pc1, fig.cap = "PC1 index loadings with confidence bands and null distribution, with filtered bootstrap analyses."}
# This plot will be assigned to a variable so it can be used later.
pc1_loadings_plot <- plot_loadings(onze_test, pc_no=1, filter_boots = TRUE)
pc1_loadings_plot 
```

In Figure \@ref(fig:filtered-pc1) we should not be surprised that the confidence
bands for <span style="font-variant: small-caps;">thought</span> F2 are high.
We have selected only those iterations of our bootstrap analysis for which this
is true. The following variables _are_ however interesting. We see that in almost
all cases with a high value for <span style="font-variant: small-caps;">thought</span> F2, 
<span style="font-variant: small-caps;">start</span> and
<span style="font-variant: small-caps;">strut</span> F2, and 
<span style="font-variant: small-caps;">thought</span> F2 take high values.
However, <span style="font-variant: small-caps;">fleece</span> F2 and <span style="font-variant: small-caps;">nurse</span> F2 can take values across a very
large range. Here, we have a clear rationale for taking the relationship between
these back vowels to be a genuine source of covariation in the data. We also,
here, attend to <span style="font-variant: small-caps;">dress</span> F2
and <span style="font-variant: small-caps;">LOT</span> F1, which appear to 
take values in the 'significant' range in almost all analyses.

If we ask which variables to _exclude_, we simply use the null distribution. 
Any sign which appears below the upper band of the null distribution should not
be used in interpretation. Of those variables which we _include_, we should be
very cautious about those whose confidence bands stretch down towards zero.

We now look at PC2. Again, we will filter our bootstrapped analyses because 
the confidence bands for the variance explained by PC1 and PC2 
overlap.
``` {r pc2-index-loadings, fig.cap = "PC2 index loadings with confidence bands and null distribution, with filtered bootstrap analyses."}
# Generate unfiltered plot for paper.
pc2_loadings_plot_uf <- plot_loadings(onze_test, pc_no=2, filter_boots = FALSE)

# Generate filtered plot to output from this block.
pc2_loadings_plot <- plot_loadings(onze_test, pc_no=2, filter_boots = TRUE)
pc2_loadings_plot 
```

Here, we again exclude those variables which fall outside the null distribution,
and note some scepticism about the inclusion of 
<span style="font-variant: small-caps;">thought</span> F2.

At this point, it is useful to look again at the variable plot to see why the
individual PCs take the form that they do.
``` {r pc1-pc2-varplot, fig.cap = "PC1 and PC2 variable plot."}
pc1_pc2_varplot <- fviz_pca_var(
  onze_pca, repel = TRUE, 
  select.var = list(contrib = 15)
)
pc1_pc2_varplot
```


Recall that the angle between arrows indicates the correlation between variables
as captured by the first two PCs, with arrows close in angle being 
highly positively correlated, arrows at right angles having no correlation,
and arrows pointing in the opposite direction being strongly negatively 
correlated. It is possible to divide this variable plot set of arrows,
mentally, into _three_ groups, rather than two. The first, the large and long
arrows for <span style="font-variant: small-caps;">thought</span>, 
<span style="font-variant: small-caps;">strut,</span> and <span style="font-variant: small-caps;">start</span>
F2, with <span style="font-variant: small-caps;">thought</span> F1, and a
smaller arrow for <span style="font-variant: small-caps;">dress</span> F2. 
This group is what tends to come through in our PC1. The second group is the 
collection of positively correlated, variables with <span style="font-variant: small-caps;">trap F1</span>
dominating. These tend to dominate PC2. The _third_ group is variously captured
by PC1 _or_ PC2, depending on the specifics of the dataset. It is dominated by
<span style="font-variant: small-caps;">fleece</span> F1 and <span style="font-variant: small-caps;">fleece</span> F2, and is correlated with <span style="font-variant: small-caps;">goose</span> F1,
<span style="font-variant: small-caps;">nurse</span> F2 etc. Across 
the range of bootstrapped analyses, this group variously
appears in either PC1 or PC2 or partially in both.^[
Consider, for instance, the high confidence band for <span style="font-variant: small-caps;">fleece</span>
F2 in Figure \@ref(fig:pc2-index-loadings).
] The two PCs we are currently
evaluating primarily place the second and third 'groups' in this variable plot
in PC2.

As always, with statistical methods, certain phenomena are brought to the fore
while others are hidden. What matters here is that we have a clear way to
discern genuine strong relationships of covariation from those which are an 
artefact of having to draw straight lines. For instance, PC1 _does_ capture
a genuine relationship of covariation between <span style="font-variant: small-caps;">thought</span>,
<span style="font-variant: small-caps;">start</span>, and <span style="font-variant: small-caps;">strut</span>.

The following is **Figure 6** in the paper.
``` {r, fig.cap = "Figure 6 for paper."}
figure_6 <- (
  (
    pc1_loadings_plot_uf + 
      labs(title = "PC1 Index Loadings", subtitle = NULL)
   ) + 
    (
      pc1_loadings_plot +
        labs(title = "PC1 Index Loadings (Filtered)", subtitle = NULL)
    )
  ) /
  (
    (
      pc2_loadings_plot_uf + 
        labs(title = "PC2 Index Loadings", subtitle = NULL)
     ) + 
    (
      pc2_loadings_plot +
        labs(title = "PC2 Loadings (Filtered)", subtitle = NULL)
    )
  ) +
  plot_annotation(
    title = "PC1 and PC2 Loadings",
    tag_levels = "A",
    theme = theme(plot.title = element_text(size = 20))
  ) +
  plot_layout(guides = "collect")

ggsave(
  here('plots', 'figure_6.png'),
  figure_6,
  dpi = 500, 
  width = 320, 
  height = 200,
  units = "mm"
)
```

We now turn to the remaining PCs which appeared as 
significant in our earlier analysis. Here we again use `filter_boots` because
PC3 and PC4 have overlapping confidence bands for variance explained.

``` {r pc3-index-loadings, fig.cap = "PC3 index loadings with confidence bands and null distribution, with filtered bootstrap analyses."}
plot_loadings(onze_test, pc_no=3, filter_boots = TRUE)
```

Here it seems that all of the significant index loadings, those from
<span style="font-variant: small-caps;">thought</span> F1 up, are really 
associated with <span style="font-variant: small-caps;">lot</span> F1. In this
case, we can recommend considering all of these variables in the interpretation.

``` {r pc4-index-loadings, fig.cap = "PC4 index loadings with confidence bands and null distribution, with filtered bootstrap analyses."}
pc4_loadings_plot <- plot_loadings(onze_test, pc_no=4, filter_boots = TRUE)
pc4_loadings_plot
```

PC4 seems to capture a correlation between <span style="font-variant:
small-caps;">strut</span> and <span style="font-variant:
small-caps;">start</span> F1, and, perhaps, <span style="font-variant:
small-caps;">kit</span> F1. This PC is not considered in Brand et al. (2021). 
It seems it is a real cross vowel relationship. However, as outlined above, 
the choice to stick with PCs which explain at least 10% of variance is
defensible in an exploratory context. Moreover, it is dominated by a single
pairwise correlation. This is not the kind of cluster which was being looked
for in Brand et al. (2021).
It remains important to report all of this, if only in supplementary material. 
It is information which may be useful for reviewers
and readers to either criticise the decisions made or as a prompt for further
research.

The following is a variable plot of PC3 and PC4, which, again, overlap in
variance.
```{r var-plot-pc3-pc4, fig.cap = "Variable plot of PC3 and PC4."}
fviz_pca_var(onze_pca, axes = c(3,4), repel = TRUE, select.var = list(contrib = 10))
```

Finally, we look at PC5. This does not overlap with any other PCs. Consequently,
we do not use the `filter_boots=TRUE` option.

``` {r pc5-index-loadings, fig.cap = "PC5 index loadings with confidence bands and null distribution, with filtered bootstrap analyses."}
pc5_loadings_plot <- plot_loadings(onze_test, pc_no=5)
pc5_loadings_plot
```

If Figure \@ref(fig:pc5-index-loadings), only <span style="font-variant:
small-caps;">dress</span> F1 appears reliably significant here. This is not a
clear pattern of covariation between vowels, so we can leave it behind. While it
might be useful for accurate data representation. It does not help with our aim
of finding covariation between vowels. We would report this as a reason for
ignoring PC5 despite it's appearing as 'significant'.

The discussion here concerns only the selection of variables with which to
interpret our PCs. We have not yet done any actual interpretation. 

**Recommendations:** We recommend excluding all variables whose index loadings
fall within the 90% interval of the null distribution. These could just as
easily be the result of random noise as signal. Where PCs are unstable, we
recommend testing the stability which obtains when bootstrapped analyses of the
data are filtered to ensure an index loading for the variable with highest
median loading above the 1st quartile. If this reveals stable structure, the
significant variables which seem to be stable should be used to interpret the
PCs. It is important to report which variables are being used to filter the
bootstraps.

## Connecting PCA with the Original Data

_This analysis corresponds to section "Returning to the data" in the paper._

Interpretation of PCs by loadings often requires both additional information and 
creative connections between the PCs and our original data.

One piece of information which is particularly useful when looking at structure
in vowel covariation is the pattern of change occurring in the vowel space over
time. Thankfully, this is easy to extract from the GAMM models we fit earlier.

Our application of PCA came after a series of modifications to the data. Now
we need to partially retrace our steps in order to understand the data. In 
both cases, we use the _scores_ assigned by PCA to each of our speakers.

There are two steps in return to our original vowel data:

1. Exploring the relationship between the data we have fed into the `prcomp`
function and the PC scores for each speaker. 
2. Exploring the vowel spaces of speakers selected by their PC score (with
the use of supplementary variables as in the 3D toy model).

We'll look at plotting change over time from GAMMs, and then consider these
two points in turn.

### Plotting Vowel Space Change

We first collect the model predictions for each gender in the data and year of
birth. We use the `get_predictions` function from `itsadug`:^[
For a tutorial on plotting vowel space changes like this see 
[here](https://joshua.wilsonblack.nz/post/visualising-vowel-space-gamms/).
]
``` {r}
# Define the values we want predictions for.
min_yob <- range(onze_vowels$yob)[[1]]
max_yob <- range(onze_vowels$yob)[[2]]

to_predict <- list(gender = c("M", "F"), yob = seq(min_yob, max_yob))

# Collect predictions for each model.
onze_preds <- onze_models %>%
  mutate(
    predictions = map(
      model, 
      ~ get_predictions(.x, cond = to_predict, print.summary=FALSE)
    )
  ) %>%
  select(
    vowel, formant_type, predictions
  ) %>%
  unnest(
    predictions
  ) %>%
  select(
    -speech_rate, -rm.ranef, -speaker
  ) %>%
  mutate(
    vowel_formant = str_c(vowel, "_", str_sub(formant_type, end=2L))
  )

onze_preds_wide <- onze_preds %>%
  select(-CI , -word) %>%
  pivot_wider(
    id_cols = c(vowel, gender, yob),
    names_from = formant_type,
    values_from = fit
  )
```

We can now visualise these:
``` {r gamm-sound-change, fig.cap = "Change in NZE monophtongs from speakers both in 1864 through 1982."}
first_obs <- onze_preds_wide %>%
  group_by(vowel, gender) %>%
  slice(which.min(yob))

sound_change_plot <- onze_preds_wide %>%
  filter(
    gender == "F"
  ) %>%
  ggplot(
    aes(
      x = F2_lob2,
      y = F1_lob2,
      colour = vowel,
      label = vowel # Add label to the aesthetics.
    )
  ) +
  geom_path(
    arrow = arrow(length = unit(3, "mm"), type = "closed"),
    show.legend = FALSE, # Remove legend
    size = 1
  ) +
  geom_label(
    # Note filtering as we are only dealing with female speakers now.
    data = first_obs %>% filter(gender == "F"), 
    show.legend = FALSE, # Again remove legend.x
    alpha = 0.7,
    size = 3
  ) +
  # Often need to use 'expansion' here to fit in 'THOUGHT'
  scale_x_reverse(expand = expansion(add = 1)) + 
  scale_y_reverse() +
  labs(
    title = "Sound Change in NZE Monophthongs",
    subtitle = "Year of birth: 1864 - 1982",
    x = "F2 (Lobanov 2.0)",
    y = "F1 (Lobanov 2.0)"
  )

sound_change_plot
```

This plot enables us to evaluate PC loadings as either:

- _all_ in or opposed to the direction of change, or
- _partially_ in the direction of change and _partially_ against it.

In the former case, the PC is plausibly interpreted as connected with sound
change.

### PCA Input and PC Scores

In any PCA analysis, especially one in which there are many steps required to
move from one to the next, it is not immediately clear what a high PC values
means with respect to the original data.

The following code uses the permutation and bootstrap test carried out by 
`pca_test` to plot the data fed to PCA (in this case, random intercepts from
GAMM models) and the PC scores assigned to our speakers. Significant variables
are indicated by colour.^[This plot is a version of one found in the
Analysis supplementary to Brand et al. (2021), in the section titled
"Relationship between intercepts and PC scores."]
We use the function `plot_pc

``` {r pca-input-and-scores, fig.dim = c(10,10), fig.cap = "PC Scores and PCA input values (random intercepts)."}
plot_pc_input(onze_pca, onze_ints %>% select(-speaker), onze_test)
```

Figure \@ref(fig:pca-input-and-scores) is useful for demonstrating that the 
PC scores assigned to individuals really do correspond to patterns in the 
data fed into PCA. For instance, it makes clear that as a speakers PC1 score
increases, their random intercept for <span style="font-variant: small-caps;">thought</span>
F1 and F2 both increase. Plots of this sort are best confined to supplementary
material.

We might also want to see what the PCs look like when represented within a 
vowel space diagram. The following code reconstructs the top right panel of
Figures 7, 9, and 11 from Brand et al. (2021).
To plot the relationship between intercepts and PC scores _within the vowel_
_space_ we add the random intercept values to the predicted fit from the GAMM
models. This spreads each speaker's random intercepts across the vowel space.
In order to plot the difference between high PC and low PC speakers, for each
vowel, we fit a series of linear models which predict the sum of the random
intercept values and the GAMM model fits (i.e. the random intercepts as
distributed across the vowel space) from the PC score. This allows us to plot
straight lines across a cloud of points for each vowel.

We first collect all of the required data together. This is a long block of
code, which is hidden by default.

``` {r, class.source = 'fold-hide'}
onze_ints_long <- onze_ints %>%
  bind_cols(onze_pca$x) %>%
  pivot_longer(
    cols = matches("F1|F2"),
    names_to = "vowel_formant",
    values_to = "intercept"
  ) %>%
  left_join(
    onze_vowels %>% select(speaker, gender, yob) %>% unique()
  ) %>%
  left_join(
    onze_preds, by = c("vowel_formant", "gender", "yob")
  ) %>%
  mutate(
    vs_intercept = intercept + fit
  )

## Linear models

onze_ints_linear <- onze_ints_long %>%
  pivot_longer(
    cols = matches("PC[0-9]+"),
    names_to = 'PC',
    values_to = 'PC_score'
  ) %>%
  group_by(vowel, formant_type, PC) %>%
  nest() %>%
  mutate(
    model = map(data, ~lm(vs_intercept ~ PC_score, data=.x)),
    data = map2(data, model, ~ .x %>% mutate(linear_fit = .y[['fitted.values']]))
  ) %>%
  select(
    data
  ) %>%
  unnest(data)


## Extract variable loadings from PCs
variable_loadings = onze_pca$rotation %>%
  as_tibble(rownames = "vowel_formant") %>%
  pivot_longer(
    cols = matches("PC[0-9]+"),
    names_to = "PC",
    values_to = "loading"
  )

## Now, add variable loadings and widen so F1 and F2 have separate columns.
onze_ints_to_plot <- onze_ints_linear %>%
  mutate(
    formant_type = str_sub(formant_type, end = 2L)
  ) %>%
  left_join(variable_loadings) %>%
  group_by(PC, vowel) %>%
  mutate(
    max_loading = max(abs(loading))
  ) %>%
  ungroup() %>%
  select(
    -fit, -CI, -intercept, -vowel_formant
  ) %>%
  pivot_wider(
    names_from = "formant_type",
    values_from = c("vs_intercept", "linear_fit", "loading"),
    names_glue = "{formant_type}_{.value}"
  )
```

The following code block defines a function which can be used to plot each PC and apply it to PC1.
``` {r pc1-vs-plot, fig.cap = "PC1 and speaker random intercepts in vowel space."}
plot_vs_pc <- function(PC_name) {
  onze_ints_to_plot %>%
      filter(
        PC == PC_name
      ) %>%
      ggplot(
        aes(
          x = F2_vs_intercept,
          y = F1_vs_intercept,
          colour = PC_score,
          group = vowel,
          label = vowel,
          size = max_loading,
          fill = rescale(max_loading^2, to = c(0, 1))
        )
      ) +
      geom_point(size = 0.7, alpha = 0.5) +
      geom_line(
        aes(
          x = F2_linear_fit,
          y = F1_linear_fit,
          size = max_loading
        ),
        show.legend = FALSE
      ) +
      geom_label_repel(
        data = . %>%
          group_by(vowel) %>%
          mutate(
            F1_vs_intercept = mean(F1_vs_intercept),
            F2_vs_intercept = mean(F2_vs_intercept)
          ) %>%
          filter(PC_score == min(PC_score)),
        show.legend = FALSE,
        alpha = 0.6,
        min.segment.length = 0,
        force = 200,
        segment.alpha = 0.5
      ) +
      scale_x_reverse(expand = expansion(add = 0.7)) +
      scale_y_reverse(position = "right") +
      scale_colour_gradient2(
        name = "PC score",
        low = "black",
        mid = "white" ,
        high = "#7CAE00",
        midpoint=0,
        breaks = c(-3, 0, 3)
      ) +
      scale_fill_gradient(low = "white", high = "yellow", guide = "none") +
      scale_size_continuous(range = c(1.5, 4), guide = "none") +
      labs(
        x = "F2 (Speaker Intercept + GAMM Fit)",
        y = "F1 (Speaker Intercept + GAMM Fit)"
      )
}

pc1_vs <- plot_vs_pc("PC1") +
  labs(
    title = "PC1 and Speaker Random Intercepts in Vowel Space"
  )
pc1_vs
```

In plots of this sort, the size and depth of yellow of the labels indicates the
size of loadings. Here, we have already decided to focus on THOUGHT, STRUT, and
START. We see that high PC scores for PC1 correspond to fronting and lowering in
<span style="font-variant: small-caps;">thought</span>, and backing in <span
style="font-variant: small-caps;">strut</span> and <span style="font-variant:
small-caps;">start</span>. This is the interpretation given in Brand et al.
(2021). This pattern is not obviously connected to changes in progress as
depicted in Figure \@ref(fig:gamm-sound-change).

``` {r pc2-vs-plot, fig.cap = "PC2 and speaker random intercepts in vowel space."}
pc2_vs <- plot_vs_pc("PC2")  +
  labs(
    title = "PC2 and Speaker Random Intercepts in Vowel Space"
  )
pc2_vs
```

PC 2, on the other hand, seems to track change in progress, particularly in
the front vowels. As PC2 scores _lower_ we see <span style="font-variant: small-caps;">trap</span>
and <span style="font-variant: small-caps;">dress</span> rising, and <span style="font-variant: small-caps;">kit</span>
lowering, just as in our 3D toy example, while changes in <span style="font-variant: small-caps;">fleece</span>
and <span style="font-variant: small-caps;">nurse</span> also match changes in progress.

``` {r pc3-vs-plot, fig.cap = "PC3 and speaker random intercepts in vowel space."}
pc3_vs <- plot_vs_pc("PC3")  +
  labs(
    title = "PC3 and Speaker Random Intercepts in Vowel Space"
  )

pc3_vs
```

This PC is interpreted in Brand et al. (2021) as capturing a relationship
between the back and front vowels (see Section 5.2.3 in Brand et al. (2021)).

The difficulty in the code above is associated with the use of 
random intercepts as input to PCA. In general, any plot of this sort needs to 
find a patch back from the input to the PCA function and speaker vowel spaces.

Another useful plot for understanding the meaning of the PCs reintroduces the
explanatory variables from our GAMM models. GAMMs were introduced to this 
project in order to ensure that the patterns of covariation. We can verify that
there is no relationship between our explanatory variables and our PC scores
by plotting them together:

``` {r pc1-by-yob-g, fig.cap = "PC1 score by year of birth and gender."}
# Create a dataframe with random intercepts, PC scores, and speaker gender and
# year of birth.
onze_ints_and_pcs <- onze_ints %>%
  bind_cols(
    onze_pca$x
  ) %>%
  left_join(
    onze_vowels %>% select(speaker, gender, yob) %>% unique()
  ) 


pc1_yob <- onze_ints_and_pcs %>%
  ggplot(
    aes(
      y = PC1,
      x = yob,
      colour = gender
    )
  ) +
  geom_point() +
  geom_smooth(method = "lm") +
  labs(
    title = "Speakers by PC1 Score, Year of Birth and Gender",
    y = "PC1 score",
    x = "Year of birth"
  ) +
  scale_colour_manual(values = c("M" = "#009E73", "F" = "#E69F00"))

pc1_yob
```

And for PC2:
``` {r pc2-by-yob-g, fig.cap = "PC2 score by year of birth and gender."}
pc2_yob <- onze_ints_and_pcs %>%
  ggplot(
    aes(
      y = PC2,
      x = yob,
      colour = gender
    )
  ) +
  geom_point() +
  geom_smooth(method = "lm") +
  labs(
    title = "Speakers by PC2 Score, Year of Birth and Gender",
    y = "PC2 score",
    x = "Year of birth"
  ) +
  scale_colour_manual(values = c("M" = "#009E73", "F" = "#E69F00"))

pc2_yob
```

In both Figure \@ref(fig:pc1-by-yob-g) and Figure \@ref(fig:pc2-by-yob-g) the
lines estimating the relationship between PC score and year of birth are flat.
This indicates that there is no obvious relationship between the two. This is
true for both genders in the data.

In the case of PC2, this is particularly important. PC2 seems to be aligned with
changes in progress over this time period. However, your PC2 score is not 
determined by your year of birth. This encourages the Brand et al. (2021) 
interpretation of PC2 as tracking how conservative a speaker is _given their year_
_of birth_. That is, the vowel space of a speaker with a low PC2 score born
in the 1890s will be quite different from that of a speaker both in the 1970s,
even while both are advanced in sound change relative to their year of birth.
We will visualise this in a moment.

The same output can easily be produced for any other PC desired. Simply change
the `x = PC2` and the plot labels from the previous code block.


The following code is to output **Figure 7** for the paper:
``` {r figure-7, fig.dim = c(10, 16), fig.cap = "All PC2 Interpretive Plots."} 
figure_7 <- pc1_yob / (pc2_vs + sound_change_plot) +
    plot_annotation(
    title = "PC2 Interpretive Plots",
    tag_levels = "A",
    theme = theme(plot.title = element_text(size = 20))
  )

ggsave(
  here('plots', 'figure_7.png'),
  figure_7,
  dpi = 500, 
  width = 320, 
  height = 300,
  units = "mm"
)

figure_7
```

### Normalised vowel spaces

One way to get back to actual vowel space data is to look at vowel spaces, 
selecting speakers by their PC scores. For this, we use the function
`plot_vowel_space` from `nzilbb.vowels`.

We start by selecting speakers who have high PC2 scores, while their other
PCs (here we include the first three PCs). By high, we mean 'in the top quartile', by
moderate, we mean 'in the second or third quartiles', by low, we mean 'in the
bottom quartile'.

``` {r}
high_pc2 <- onze_ints_and_pcs %>%
  filter(
    PC2 > quantile(PC2, 0.75),
    if_all(
      .cols = c("PC1", "PC3"), 
      ~ between(.x, quantile(.x, 0.25), quantile(.x, 0.75))
    )
)

high_pc2_speakers <- high_pc2 %>% pull(speaker)
high_pc2_speakers
```
We plot these speakers using `plot_vowel_space`. This function requires the
data to have columns in the order "speaker, vowel identifiers, F1, and F2".
Here, I will modify the speaker identifiers so they include year of birth as
well. This is convenient for our interpretation of PC2 with respect to change
in progress.
``` {r high-pc2-vs, fig.dim = c(10,10), fig.cap = "High PC2 Vowel Spaces."}
# We define some vowel colours for the NZE monophthongs.
vowel_colours <- c(
  DRESS = "#9590FF",
  FLEECE = "#D89000",
  GOOSE = "#A3A500",
  KIT = "#39B600",
  LOT = "#00BF7D",
  NURSE = "#00BFC4",
  START = "#00B0F6",
  STRUT = "#F8766D",
  THOUGHT = "#E76BF3",
  TRAP = "#FF62BC"
)

plot_vowel_space(
  onze_vowels %>% 
    select(speaker, vowel, F1_lob2, F2_lob2, yob) %>%
    mutate(
      speaker = paste0(speaker, " (", yob, ")")
    ), 
  speakers = glue("{high_pc2$speaker} ({high_pc2$yob})"), 
  vowel_colours = vowel_colours
) + labs(
  title = "High PC2 Score Speakers"
)
```

We can do the same for low PC2 speakers:
``` {r low-pc2-vs, fig.dim = c(10,10), fig.cap = "High PC2 Vowel Spaces."}
low_pc2 <- onze_ints_and_pcs %>%
  filter(
    PC2 < quantile(PC2, 0.25),
    if_all(
      .cols = c("PC1", "PC3"), 
      ~ between(.x, quantile(.x, 0.25), quantile(.x, 0.75))
    )
)

plot_vowel_space(
  onze_vowels %>% 
    select(speaker, vowel, F1_lob2, F2_lob2, yob) %>%
    mutate(
      speaker = paste0(speaker, " (", yob, ")")
    ), 
  speakers = glue("{low_pc2$speaker} ({low_pc2$yob})"), 
  vowel_colours = vowel_colours
) + labs(
  title = "Low PC2 Score Speakers"
)
```

And, finally, we pick some entirely moderate speakers. That is, speakers who
are in the middle for the first three PCs. We will tighten our bands so that these
speakers are 'super moderate', in the sense that they all fall within the 0.3 and
0.7 quantiles.
``` {r moderate-vs, fig.dim = c(10,20), fig.cap = "Moderate speaker vowel spaces."}
moderates <- onze_ints_and_pcs %>%
  filter(
    if_all(
      .cols = c("PC1", "PC2", "PC3"), 
      ~ between(.x, quantile(.x, 0.3), quantile(.x, 0.7))
    )
)

plot_vowel_space(
  onze_vowels %>% 
    select(speaker, vowel, F1_lob2, F2_lob2, yob) %>%
    mutate(
      speaker = paste0(speaker, " (", yob, ")")
    ), 
  speakers = glue("{moderates$speaker} ({moderates$yob})"), 
  vowel_colours = vowel_colours
) + labs(
  title = "Moderate Speakers"
)
```

There is no short cut to insight with plots like this. In the following section, 
we will pick out four speakers to illustrate the interpretation of PC2.

Sometimes it is useful to produce an interactive in order to explore the vowel
spaces of individual speakers by PC score. This was done as part of Brand et al.
(2021), and can be found [here](https://onze.shinyapps.io/Covariation_shiny/).

### Interpretation: bringing it all together

All of this information can be brought together in composite plots. We do this,
as an example, for PC2. 

We combine plots using the `patchwork` library, which enables us to apply
arithmetic to `ggplot` objects. For instance, we can put two plots side by
side by joining them with a `+`.

Here we show the difference between speakers with moderate PC2 from early
and late in our time period and two speakers from the middle of the time 
period with extreme scores. 

From the moderate speakers (Figure \@ref(fig:moderate-vs)) we pick the speaker
`MU_f_463`, although we could just have easily have picked `MU_m_072`. This
speaker will represent an early born speaker with moderate PCs. We will pick
`CC_f_130` as a late born moderate speaker. We then pick `CC_f_457` as a 
high PC2 speaker, and `CC_f_545` as a low PC2 speaker from around the same year.

We first highlight these in our plot of PC2 scores by year and gender.

``` {r pc2-patch-1, fig.cap = "PC2 score by year of birth and gender."}
desired_speakers <- c("MU_f_463", "CC_f_130", "CC_f_457", "CC_f_545")

pc2_patch_1 <- onze_ints_and_pcs %>%
  mutate(
    desired_speaker = speaker %in% desired_speakers,
  ) %>%
  ggplot(
    aes(
      y = PC2,
      x = yob,
      colour = gender,
      label = speaker
    )
  ) +
  geom_point(aes(alpha = desired_speaker), show.legend = FALSE) +
  geom_smooth(method = "lm", alpha = 0.5) +
  geom_label_repel(
    data = . %>% filter(speaker %in% desired_speakers),
    min.segment.length = 0
  ) +
  labs(
    title = "Speakers by PC2 Score, Year of Birth and Gender",
    y = "PC2 score",
    x = "Year of birth"
  ) +
  scale_colour_manual(values = c("M" = "#009E73", "F" = "#E69F00")) +
  scale_alpha_manual(values = c("TRUE" = 1, "FALSE" = 0.2))

pc2_patch_1
```

We can also combine this plot with a plot of each speakers vowel space as 
follows (This is **Figure 8** in the paper):
``` {r fig.dim = c(8, 16)}
figure_8 <- pc2_patch_1 /
  plot_vowel_space(
    onze_vowels %>%
      select(speaker, vowel, F1_lob2, F2_lob2, yob),
    speakers = desired_speakers,
    vowel_colours = vowel_colours
  ) +
  plot_annotation(
    title = "Speakers by PC2 Score with Vowel Spaces",
    tag_levels = "A",
    theme = theme(plot.title = element_text(size = 20))
  ) +
  plot_layout(heights = c(2, 4))

ggsave(
  here('plots', 'figure_8.png'),
  figure_8,
  dpi = 500, 
  width = 200, 
  height = 300,
  units = "mm"
)


figure_8
```

This plot is discussed in more detail in the paper. The main point is to 
highlight that speakers _born in the same year_ can be identified as leaders
or laggers by PC2 _and_ that this difference can be seen in the speakers 
vowel spaces. We also see how moderate speakers in PC2 have very different
looking vowel spaces if selected from the start or the end of the year of birth
range.

# Conclusion

This RMarkdown has set out the use of PCA for monophthong data in two stages.
Both have used data from the ONZE corpus. The first stage presented the 
basic idea of PCA using three variables from the ONZE dataset with known 
inter-relations. The aim of this section was to present PCA geometrically, 
visualising the centre of a cloud of points, and how the PCs are generated 
by drawing lines which successively maximise variance.

The second stage presented a simplified, standalone, version of our case study:
the PCA analysis of the ONZE data carried out by Brand et al. (2021). We fit a
series of GAMM models from which we extract random intercepts for each speaker.
This enabled us to remove known variation from our data, to further enhance our
ability to find new patterns of variation. We then applied PCA to the collection
of random intercepts, first showing how to fit and interact with a PCA analysis
in R before turning to three core questions:

1. When is PCA appropriate?
2. How many PCs should be considered?
3. Which variables should we use to interpret the PCs?

The approach to these questions taken in Brand et al. (2021) was extended so as
to be more robust to instability in PC loadings and includes a more objective
method for _excluding_ loadings. Recommendations were given for responding to
each question.

